### 2022-02-10 chapter1/chapter.tex
.

%!TEX root = ../thesis.tex
\chapter{Introduction \& Preliminaries \label{ch:intro}} 
% THE ROUGH IDEA & CONTEXT
% 	THE CONTEXT: FLUID-FLUID QUEUES
%	PDEs 
%	NON-NEGATIVITY & CONSERVATION 
% MATHEMATICAL PRELIMINARIES
A fluid queue is a two-dimensional stochastic process \(\{\bs X(t)\} = \{(X(t),\varphi(t))\}_{t\geq0}\). The phase process, also known as the driving process, \(\{\varphi(t)\}_{t\geq0}\), is a continuous-time Markov chain (CTMC). The level process, \(\{X(t)\}_{t\geq0}\), is a real-valued, continuous, and piecewise linear.

Stochastic fluid queues have found a variety of applications such as telecommunications (see \cite{anick1982} as a canonical application in this area), power systems \citep{hydro}, risk processes \citep{betal2005} and environmental modelling \citep{wurm2020}. Fluid queues are relatively well studied. Largely, the analysis of fluid queues falls into two categories, matrix-analytic methods \citep{ajr2005,ar2003,ar2004,bean2005b,bean2005,bot08,bean2009,dasilva2005,latouche2018}, and differential equation-based methods \citep{anick1982,kk1995,beanetal2019}. %For example, Ramaswami CITE, analysed fluid queues by mapping them to a quasi-birth-and-death process (QBD), after which they applied known matrix-analytic methods for QBDs to compute quantities of interest. Anick Mitra Sondhi CITE, analysed fluid queues using a more direct differential equation-based method. Since Ramaswami's CITE initial work, there has been significant developments in the analysis of fluid queues CITE and related algorithms CITE.

More recently, \cite{bo2014} extended fluid queues to \emph{so-called} stochastic fluid-fluid queues. In a fluid-fluid queue there is a second level process, \(\{Y(t)\}_{t\geq0}\) which is itself driven by a fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). The analysis,  \citep{bo2014}, is in principal similar to the matrix-analytic methods of \citep{bean2005}, and derives results about the second level process \(\{Y(t)\}_{t\geq0}\) in terms of the infinitesimal generator (a differential operator) of the fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). For practical computation of the results of \cite{bo2014}, a matrix-discretisation of the infinitesimal generator of the fluid queue can be used. To this end, to date, two possible discretisation have been suggested. Taking a differential equations-based approach, \cite{beanetal2019} use the discontinuous Galerkin (DG) method to discretise this operator, while \cite{bo2013} take a stochastic modelling and matrix-analytic methods approach to approximate the fluid-queue by a quasi-birth-and-death (QBD) process. Both approaches are insightful and offer different tools and perspectives with which to analyse the resulting approximations. It turns out that the latter approach is a sub-class of the former; the QBD can be viewed as the simplest DG scheme where the operator is projected onto a basis of piecewise constant functions.

%A QBD can be viewed as a two-dimensional CTMC, \(\{(L(t),\varphi(t))\}_{t\geq0}\), where \(\{L(t)\}\) is the discrete level process, and \(\{\varphi(t)\}_{t\geq0}\) is the phase process. The level process \(\{L(t)\}\) is skip free, meaning that, given the process is at level \(L(t)=\ell\), is may only jump to \(\ell+1\) or \(\ell-1\) at jump epochs. The sojourn time of \(\{L(t)\}\) in a given level follows a phase-type distribution

In the context of approximating fluid queues, one advantage of the QBD discretisation and, equivalently, DG schemes with constant basis functions, is that they guarantee probabilities computed from the approximation are positive \citep[Section 3.3]{koltai2011}, see also \citep{nodalDGBook} and references therein. One justification for the positivity preserving property is from the interpretation of the discretisation as a stochastic process thereby ensuring positivity. For higher order DG schemes there is no such interpretation. Moreover, higher-order DG approximation schemes may produce negative, or highly oscillatory solutions, particularly when discontinuities or steep gradients are present. Methods to navigate the problem of negative or highly oscillatory solutions have been developed, such as slope limiters, and filtering (see \citep[Section 6.5]{nodalDGBook} and references therein). Slope limiting effectively alters the discretised operator in regions where oscillations are detected and lowers the order of the approximation in theses regions. Filtering is a post-hoc method which looks to recover an accurate solution, given an oscillatory approximation.

\section{Fluid queues}
A fluid queue is a two-dimensional stochastic process \(\{\bs X(t)\} = \{(X(t),\varphi(t))\}_{t\geq0}\) where \(\{\varphi(t)\}_{t\geq0}\) is known as the phase or driving process, and \(\{X(t)\}_{t\geq0}\) is known as the level process or buffer. The phase process \(\{\varphi(t)\}_{t\geq0}\), is an irreducible continuous-time Markov chain (CTMC) with finite state space, which we we assume to be \(\mathcal S=\{1,2,\dots,N\}\) without loss of generality, and infinitesimal generator \(\bs T= [T_{ij}]_{i,j\in\mathcal S}\). We assume that \(\bs T\) is \emph{conservative}. Associated with states \(i\in\mathcal S\) are real-valued \emph{rates} \(c_i\in\mathbb R\).

Partition the state space \(\calS\) into \(\calS_+ = \{i\in\calS\mid c_i>0\}\), \(\calS_- = \{i\in\calS\mid c_i<0\}\) and \(\calS_0 = \{i\in\calS\mid c_i=0\}\). We assume, without loss of generality, that the generator \(\bs T\) is partitioned into sub-matrices 
\[\bs T = \left[\begin{array}{ccc}\bs T_{++} & \bs T_{+-} & \bs T_{+0} \\ \bs T_{-+} & \bs T_{--} & \bs T_{-0} \\ \bs T_{0+} & \bs T_{0-} & \bs T_{00}  \end{array}\right],\]
where \(\bs T_{mn} = [T_{ij}]_{i\in\mathcal S_m, j\in\mathcal S_n}\), \(m,n\in\{+,-,0\}\). Also define the diagonal matrices 
\begin{align*}
	\bs C &= \left[\begin{array}{ccc} \bs C_+ && \\ &\bs C_-& \\ && \bs 0\end{array}\right], && \bs C_+ = diag(c_i,i\in\calS_+), && \bs C_- = diag(|c_i|,i\in\calS_-),
\end{align*}
where \(diag(a_i,i\in\mathcal I)\) denotes a diagonal matrix with entries \(a_i\) down the diagonal.

When no boundary conditions are imposed, the level process is given by 
\[X(t) = X(0) + \int_{s=0}^t c_{\varphi(s)}\wrt s.\]
Sample paths of \(\{X(t)\}\) are continuous and piecewise linear, with \(\cfrac{\wrt }{\wrt t} X(t) = c_\varphi(t)\), when \(X(t)\) is differentiable. Given sample paths of \(\{\varphi(t)\}\), then \(\{X(t)\}\) is deterministic, and in this sense, \(\{\varphi(t)\}\) is the only stochastic element of the fluid queue.

Often, boundary conditions are imposed. Here, we consider a mixture of \emph{regulated} and \emph{reflecting} boundary conditions. Upon hitting a boundary we suppose that, with probability \(p_{ij},\,i,j\in\mathcal S\), the phase process instantaneously transitions from phase \(i\) to phase \(j\) (note that we might have \(i=j\) i.e.~no transition). At a lower boundary, if \(j\in\calS_0\cup\calS_-\), then \(\cfrac{\wrt}{\wrt t} X(t) = 0\), and the phase process continues to evolve according to the sub-generator 
\[\left[\begin{array}{cc} \bs T_{--} & \bs T_{-0} \\ \bs T_{0-} & \bs T_{00}  \end{array}\right],\]
until such a time that \(\varphi(t)\) transitions to a phase \(k\in\calS_+\), at which time \(X(t)\) leaves the boundary. Similarly, at an upper boundary if \(j\in\calS_0\cup\calS_+\), then \(\cfrac{\wrt}{\wrt t} X(t) = 0\) and the phase process continues to evolve according to the sub-generator 
\[\left[\begin{array}{cc} \bs T_{++} & \bs T_{+0} \\ \bs T_{0+} & \bs T_{00}  \end{array}\right],\]
until such a time that \(\varphi(t)\) transitions to a phase \(k\in\calS_-\) at which time \(X(t)\) leaves the boundary. Without loss of generality, we assume the lower and upper boundaries (when present) are at \(x=0\) and \(x=M>0\), respectively.

In summary, the evolution of the level can be expressed as 
\[\cfrac{\wrt}{\wrt t} X(t) = \begin{cases} c_{\varphi(t)}, & \mbox{ if } X(t)>0, \\ \max\{0,c_{\varphi(t)}\}, & \mbox{ if } X(t)=0, \\ \min\{0,c_{\varphi(t)}\}, & \mbox{ if } X(t)=M.  \end{cases}\]

Let \(\bs f(x,t) = \vligne{f_i(x,t)}_{i\in\calS}\) be a row-vector function where \(f_i(x,t)\) is the density of \(\mathbb P(X(t)\leq x, \varphi(t) = i)\), assuming it exists. When a differentiable density exists, the system of partial differential equation which describes the evolution of the densities \(\bs f(x,t)\) is 
\begin{equation}
	\cfrac{\partial}{\partial t} \bs f(x,t) = \bs f(x,t)\bs T - \cfrac{\partial}{\partial x}\bs f(x,t)\bs C.\label{eqn: pde}
\end{equation}
The initial condition is the initial distribution of the fluid queue, \(f_i(x,0)\). Often a differentiable density function does not exist and therefore the partial differential equation (\ref{eqn: pde}) is not well-defined. For example, for a fluid queue with a regulated boundary, if the initial distribution of the fluid queue is a point mass at any point \(x_0\geq 0\) and in phase \(i\in\calS_+\cup\calS_0\), then a density function \(f_i(x,t)\) will not exist for any finite \(t\). Specifically, a point mass will persist along the ray \(x_0+c_it\), \(t\geq 0\). In such situations, it is the \emph{weak solution} to (\ref{eqn: pde}) that we seek. A weak solution satisfies
\begin{equation}
	-\int_x\int_t \bs f(x,t)\cfrac{\partial}{\partial t} \bs \psi(x,t)\wrt t \wrt x= \int_{x}\int_t\bs f(x,t)\bs T\bs \psi(x,t)\wrt t \wrt x + \int_x\int_t\bs f(x,t)\bs C \cfrac{\partial}{\partial x} \bs psi(x,t)\wrt t \wrt x,\label{eqn: weak pde}
\end{equation}
for every vector of test functions, \(\bs \psi(x,t)\), wich are smooth and have compact support.

Boundary conditions may also be imposed on (\ref{eqn: pde}).

Discretisation methods approximate the operator on the right-hand side of (\ref{eqn: pde}) by a matrix, in our case, by the generator of a QBD-RAP.

\section{Matrix-exponential distributions}
Here we recount some facts about matrix exponential distributions. See \citep{MEinAP} for a more detailed exposition. A random variable, \(Z\), is said to have a matrix-exponential distribution if it has a distribution function of the form \(1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s\), where \(\bs \alpha\) is a \(1\times p\) \emph{initial vector}, \(\bs S\) a \(p\times p\) matrix, and \(\bs s\) a \(p\times 1\) \emph{closing vector}, and \(\displaystyle e^{\bs S x} := \sum_{n=0}^\infty \cfrac{\left(\bs Sx\right)^n}{n!}\) is the matrix exponential. The density function of \(Z\) is given by \(f_Z(x) = \bs\alpha e^{\bs S x}\bs s\). The only restrictions on the parameters \((\bs \alpha, \bs S, \bs s)\) are that \(\bs \alpha e^{\bs S x} \bs s\) be a valid density function, i.e.~\(\bs \alpha e^{\bs S x} \bs s\geq 0,\) for all \(x\geq 0\) and \(\lim_{x\to\infty} 1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s = 1.\) There is the possibility of an \emph{atom} (a point mass) at 0, but here we do not consider this possibility. These condition that \(\bs \alpha e^{\bs Sx}\bs s\) be a valid density imposes some properties on representations \((\bs \alpha, \bs S, \bs s)\). However, in general there is no way to determine whether, given a triplet \((\bs \alpha, \bs S, \bs s)\) whether it is a representation of a matrix-exponential distribution, or not. Nonetheless, some properties of a triplet \((\bs \alpha, \bs S, \bs s)\) are known, such as the following, which is used in the characterisation of QBD-RAPs. 
\begin{thm}[Theorem 4.1.3, \cite{MEinAP}]
	The density function of a matrix-exponential distribution with representation \((\bs\alpha, \bs S, \bs s)\) can be expressed in terms of real-valued constants as 
	\begin{align}
		\psi(x)&=\sum_{j=1}^{m_1} \sum_{k=1}^{p_j} c_{jk} \cfrac{x^{k-1}}{(k-1)!} e^{\mu_jx} + \sum_{j=1}^{m+2}\sum_{k=1}^{q_j} d_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\cos(\sigma_jx) \nonumber 
		\\&\quad{}+ \sum_{j=1}^{m_2}\sum_{k=1}^{q_j}e_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\sin(\sigma_jx), \label{eqn: spec}
	\end{align}
	for integers \(m_1,\,m_2,\,p_j,\) and \(q_j\) and some real constants \(c_{jk},\,d_{jk},\,e_{jk},\,\mu_j,\,\eta_j,\) and \(\sigma_j\). Here \(\mu_j,\, j=1,\dots,m_1\) are the real eigenvalues of \(\bs S\), while \(\eta_j+i\sigma_j, \, \eta_j-i\sigma_j,\, j=1,\dots,m_2\) denote its complex eigenvalues, which come in conjugate pairs. Thus \(m_1+2m_2\) is the total number of eigenvalues, while the dimension of the representation is given by \(\displaystyle p=\sum_{j=1}^{m+1}p_j + 2\sum_{j=1}^{m_2}q_j\). 
\end{thm}
\begin{thm}[Theorem 4.1.4, \cite{MEinAP}]\label{thm: 4.1.4}
	Consider the nonvanishing terms of the matrix exponential density (\ref{eqn: spec}), \emph{i.e.}, the terms for which \(c_{jk}=0,\,d_{jk} =0,\) or \(e_{jk}=0\). Among the corresponding eigenvalues \(\lambda_j\), there is a real dominating eigenvalue \(\kappa\), say. That is, \(\kappa\) is real, \(\kappa \geq Re(\lambda_j)\) for all \(j\), and the multiplicity of \(\kappa\) is at least the multiplicity of every other eigenvalue with real part \(\kappa\).
\end{thm}
\begin{cor}[Corollary 4.1.5, \cite{MEinAP}]
If \((\bs \alpha, \bs S, \bs s)\) is a representation for a matrix-exponential distribution, then \(\bs S\) has a real dominating eigenvalues.
\end{cor}
\begin{thm}[Theorem 4.1.6, \cite{MEinAP}]
	Let \(Z\) be a matrix-exponentially distributed random variable with density (\ref{eqn: spec}). Then the dominant real eigenvalue \(\kappa\) of Theorem~\ref{thm: 4.1.4} is strictly negative. 
\end{thm}
We define \(dev(\bs S)\) to be the real dominating eigenvalue of \(\bs S\), that is \(dev(\bs S)=\kappa\) in Theorem~\ref{thm: 4.1.4}.

>>>>

%!TEX root = ../thesis.tex
\chapter{The Existing Literature \& Mathematical Preliminaries \label{ch:math}} 
% intro to intro
% contextual background
	% Fluid queues 
		% definition
		% applications 
		% existing results 
	% Fluid-fluid queues
		% definition 
		% basic analysis in terms of generators
		% the need for approximation and how the old work doesn't fit here
	% The DG method
		% pertition into cells
		% project on to a basis
		% problems: oscillations
		% soutions and why they dont fit
		% observation about constant-basis methods
		% this thesis asks the question: motivated by the constant basis being a probability model, can we dream up a more accurate probability model to approximate a fluid queue?
	% On the structure of the constant basis/uniformisation method (QBD)
	% least variable is PH, so what about MEs?
% structure of the thesis
	% rest of chapter 1 gives mathematical preliminaries
	% chapter 2 explains the DG method, problems/oscillations, slope limiting, and appllication to SFFMs
	% inspired by the structure of the order-1 scheme, and to solve the negativity problem, chapter 3 introduces a new approximation method; The QBD-RAP. In this chapter we derive the intuitively of the model and explain it's dynamics.
	% we then move to convergence, with chapter 4 proving a convergence of the QBD-RAP up to the first hitting time of the fluid queue on the edges of an interval. technical results and extensions are left to the appendix: certain properties of closing vectors, convergence without ephemeral phases, and certain matrix algebraic manipulations. 
	% chapter 5 stitches together the results of chapter 4 to prove a global result about convergence
	% chapter 6 investigate, numerically, the performance of the DG, DG with limiter, uniformisation and QBD-RAP schemes. 
	% chapter 7  makes concluding remarks.
% mathematical preliminaries 
	% CTMCs
	% fluid queues
	% fluid-fluid queues and operator-analytic expressions (all of them)
		% the equations of bo2014
		% meets the partition of the DG/QBD-RAP, and reconstructing the Fil sets 
    % why we cant solve the equations, what we need to approximate to solve it, i.e. B, R, then D, Psi
	% projections
	% phase-type distributions 
		% definition
		% least variable property
	% matrix exponentials
		% definition 
		% properties 
		% verification that parameters give an ME
		% CMEs
	% QBD-RAPs, then QBDs as a special case
		% orbit processes, their interpretation for PH and how they differ for MEs
	% convergence theorems 
		% DCT 
		% Continuity of Laplace transforms

% THE ROUGH IDEA & CONTEXT
% 	THE CONTEXT: FLUID-FLUID QUEUES
%	PDEs 
%	NON-NEGATIVITY & CONSERVATION 
% MATHEMATICAL PRELIMINARIES
\section{Stochastic processes}
% \subsubsection{Random variables}
% Let \(\mathcal S\) be a set.

% Roughly following \cite{billingsleyconvergence}, a random variable \(Z\), is a mapping between a \emph{sample space}, \(\mathcal S\), say, to the real numbers.

% An \emph{event} is a set \(\{\omega \in \mathcal S \mid Z(\omega)\in E\}\equiv \{\omega\in Z^{-1}(E)\}\) where \(E\subseteq \mathcal S\) and \(Z^{-1}\) is the inverse image of the mapping \(Z\). Associated with a \(Z\) is a \emph{probability measure}, \(P\), which maps events to \([0,1]\) and has the properties \(P(\emptyset)=0\), \(P(S)=1\) and if . The notation \(P(\{\omega\in Z^{-1}(E)\})\) is cumbersome, so we write \(\mathbb P(Z\in E)\) in its place.

% The sample space, \(\mathcal S\), is the set of all possible outcomes. Associated with the random variable \(Z\) is a \emph{probability measure} which assigns

A stochastic process is a sequence, \(\{X(t)\}_{t\in\mathcal T}\), of random variables or random vectors indexed by some index set \(\mathcal T\). In this thesis in the case that the index set is \(\mathcal T=\{0,1,2,...\}\) we say that the process \(\{X(n)\}_{n\in \mathcal T}\) is a discrete-time process, and we will typically use the dummy variable \(n\). When the index set is \(\mathcal T=[0,\infty)\) we say write \(\{X(t)\}_{t\in\mathcal T}\) is a continuous-time process, and we will typically use the dummy variable \(t\). We may omit the index set and write \(\{X(t)\}\) in place of \(\{X(t)\}_{t\in\mathcal T}\) when it is not explicitly needed, or we may write \(\{X(t)\}_{t\geq t_0}\) to mean \(\{X(t)\}_{t\in[t_0,\infty)}\). The \emph{sample space} of \(X(t)\) is the set of possible values that \(X(t)\) can take at any \(t\in\mathcal T\).

The \emph{initial distribution}, \(\mu\), of a stochastic process is the distribution of \(X(0)\). More generally, for a random variable, \(Z\), we write \(Z\sim \nu\) when \(Z\) has the distribution given by the probability measure \(\nu\). For the probability that \(X(t)\) lies in some set \(E\subset \mathcal S\) we write \(\mathbb P(X(t)\in E \mid X(0)\sim \mu)\). When \(\mu\) assigns probability 1 to a single point, \(x\in\calS\), say, we write \(\mathbb P(X(t)\in E \mid X(0)=x).\) A stochastic process is \emph{stationary} if
\begin{align}\label{eqn: stationary}
	&\nonumber\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	\\&= \mathbb P(X(t_n+t)=i,X(t_{n-1}+t)\in E_{n-1},...,X(t_0+t)\in E_0),
\end{align}
for any \(t,t_0,...,t_n\in\mathcal T\) and any \(E_0,...,E_n\subseteq \mathcal S\).
A random variable, \(\tau\), is a \emph{stopping time} for the stochastic process \(\{X(t)\}_{t\in\mathcal T}\) if \(\tau\in\sigma(X(s), s\leq t),\) where \(\sigma(X(s), s\leq t)\) denotes the \(\sigma\)-algebra generated by \(\{X(s), s\leq t\}\).

We call \(\{X(n)\}_{n\in\{0,1,2,\dots\}}\) a \emph{discrete-time Markov chain} if \(\calS\) is countable
\begin{align}\label{eqn: markov property}
	\nonumber &\mathbb P(X(n+1)=j \mid X(n)=i,X(n-1)=i_{n-1},...,X(0)=i_0) 
	\\&= \mathbb P(X(n+1)=j \mid X(n)=i) 
\end{align}
for all \(n\in\{0,1,2,...\}\) and \(i,i_0,...,i_{n-1},j\in\calS,\) and refer to (\ref{eqn: markov property}) as the Markov property. The process \(\{X(n)\}_{n\in\{0,1,2,...\}}\) is said to be time-homogeneous if 
\(\mathbb P(X(n+1)=j \mid X(n)=i)\) does not depend on \(n\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(n)\}_{n\in\{0,1,2,...\}}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong markov property}
	\mathbb P(X(\tau+1)=j \mid \sigma(X(0),X(1),...,X(\tau))) = \mathbb P(X(\tau+1)=j \mid X(\tau)).
\end{align}

We call \(\{X(t)\}_{t\geq 0}\) a \emph{continuous-time Markov chain} if \(\calS\) is countable and for all \(t_{n+1} > t_n > t_{n-1} > ... > t_0 > 0\) and \(i,i_0,...,i_{n-1},j\in\calS,\)
\begin{align}\label{eqn: markov property cmtc}
	\mathbb P(X(t_{n+1})=j \mid X(n)=i, X(n-1)=i_{n-1},...,X(0)=i_0) = \mathbb P(X(t_{n+1})=j \mid X(t_n)=i).
\end{align}
The process \(\{X(t)\}_{t\geq 0}\) is said to be time-homogeneous if 
\(\mathbb P(X(t+s)=j \mid X(s)=i)\) does not depend on \(s\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(t)\}_{t\geq 0}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong markov property ctmc}
	\mathbb P(X(\tau + t)=j \mid \sigma(X(s),s\leq \tau)) = \mathbb P(X(\tau+t)=j \mid X(\tau)).
\end{align}

Most generally, we call \(\{X(t)\}_{t\geq 0}\) a \emph{Markov process} if for all \(t_n > t_{n-1} > ... > t_0 > 0,\, t_k\in\mathcal T,\, k=0,...,n+1,\) and \(E_0,...,E_n\subseteq \mathcal S\)
\begin{align}\label{eqn: markov process}
	\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	= \mathbb P(X(t_n)\in E_n\mid X(t_{n-1})),
\end{align}
and is time homogeneous if \(\mathbb P(X(t+s)\in E_n\mid X(s))\) does not depend on \(s\).

The Chapman-Kolmogorov equations state that 
\begin{align}\label{eqn: CK}
	&\nonumber \mathbb P(X(t+s)\in E\mid X(0)\sim \mu) 
	\\&= \int_{x\in\calS}\mathbb P(X(t+s)\in E\mid X(t)=x)\mathbb P(X(t)\in \wrt x\mid X(0)\sim \mu).
\end{align}
% Markov processes, time-homogeneity, strong markov prperty, partitioning and LOTP

\section{Some semigroup theory}
The evolution of Markov processes can be described by an \emph{operator semigroup}. Furthermore, semigroups arise in other contexts related to the analysis of fluid queues and fluid-fluid queues. Semigroups (and therefore Markov processes) can be characterised by their \emph{infinitesimal generator}, or generator for short. One of the aims of this thesis is to approximate the generator of a type of Markov process called a fluid queue, which we introduce more formally in Section~\ref{sec: fqs}. Here we very briefly introduce semigroups and their \emph{infinitesimal generator}.

\subsubsection{Operators}
Let \(L\) be a Banach space (a vector space with a metric and where Cauchy sequences converge to a point within the space). An operator is a mapping which takes elements of a vector space and sends them to elements of a vector space (not necessarily the same space). That is, let \(V,W\) be normed vector spaces, then \(\mathbb B:\mathcal D(\mathbb B)\subset V\to W\) is an operator. The set \(\mathcal D(\mathbb B)\) is the domain of the operator, and is not necessarily all of \(L\), when specification of the domain is unnecessary we sometimes write \(\mathbb B:V\to W\). Let \(c\in \mathbb R\) and \(v\in V\), \(w\in W\). Then \(\mathbb B\) is a linear operator if \(\mathbb B(cv) = c(\mathbb Bv)\) and \(\mathbb B(v+w)=\mathbb Bv+\mathbb Bw\). We define the norm of an operator as \(||\mathbb B||=\sup_{v\in V}\cfrac{||\mathbb Bv||_W}{||v||_V}\), where \(||\cdot||_V\) and \(||\cdot||_W\) are the norms attached to \(V\) and \(W\) respectively.  An operator is said to be bounded if \(||\mathbb B||<M\) for some \(\infty > M\in \mathbb R\), for ever \(v\in V\).

% For example, let \(L\) be the space of real, smooth functions for which \(f(x)\to 0\) as \(x\to \infty\), and consider the sup norm, \(||f||=\sup\limits_{x\in\mathbb R}f(x)\). Then differentiation is an operator which maps \(L\) to it self. That is \(B=\cfrac{\wrt}{\wrt x}:L\to L\). In addition \(B\) is linear since for any \(f,g\in L\), \(B(f+g) = \cfrac{\wrt}{\wrt x}(f(x)+g(x)) =  \cfrac{\wrt}{\wrt x}f(x) +  \cfrac{\wrt}{\wrt x}g(x)\) and \(B(cf)= \cfrac{\wrt}{\wrt x}cf(x)=c \cfrac{\wrt}{\wrt x}f(x)\). The operator \(B\) is, however, not bounded.

% The \emph{graph} of \(\mathbb B\) is defined as the linear subspace of \(L\times L\), \(\mathcal G(B)=\{(f,Bf):f\in \mathcal D(B)\}\subset V\times W\}\). The operator \(B\) is said to be closed if \(\mathcal G(B)\) is a closed subset of \(V\times W\). That is, for every sequence \(\{f_n\}\) in \(\mathcal D(B)\) such that \(f_n\to f\) and \(Bf_n=g_n\to g \), then \(f\in\mathcal D(B)\) and \(Bf = g\). Note that we must have both \(\{f_n\}\) and \(\{g_n\}\) converge.

% For example, let \(B=\cfrac{\wrt}{\wrt x}:C^1[0,1]\subset C^0[0,1]\to C^0[0,1]\), where \(C^n[0,1]\) is the set of all \(n\)-times continuously differentiable functions on the interval \([0,1]\). The graph of \(B\) is the set \(G(B)=\{(f,Bf); f\in C^1[0,1]\}\). The operator \(B\) is closed since for any \(f_n\in C^1[0,1]\) with \(f_n\to f\) and \(Bf_n = \cfrac{\wrt f_n}{\wrt x}=g_n\to g\), then \(f\in C^1[0,1]=\mathcal D(B)\) and \(Bf=g\). Notice that we can construct \(\{f_n\}\) such that \(f_n\to f\notin \mathcal D(B)\) (e.g.~from the Weierstrass Approximation Theorem there is a sequence of polynomials in \(C^1[0,1]\) converging to \(f(x)=|x-1/2|\notin C^1[0,1]\)) but in these cases \(Bf_n = \cfrac{\wrt f_n}{\wrt x}=g_n\) must not converge.

\subsubsection{Semigroups}  A family of linear operators is a collection of indexed linear operators \(\{\mathbb B(i):i\in I\}\) where \(I\) is some index set. That is, for each \(i\in I\), \(\mathbb B(i)\) is a linear operator. A family of operators \(\{T(t):L\to L:t\geq 0\}\) is said to have the \emph{semigroup} property if \(\mathbb T(t+s)=\mathbb T(t)T(s)\). If, furthermore, \(\mathbb T(0)=I\), the identity operator, then the family \(\{\mathbb T(t):t\geq 0\}\) is known as a semigroup. If \(\mathbb T(t)f\to f\) as \(t\to 0^+\), then \(\{\mathbb T(t):L\to L:t\geq 0\}\) is said to be strongly continuous. If the \(||\mathbb T(t)||\leq 1\) for all \(t\geq 0\) then \(\{\mathbb T(t):L\to L:t\geq 0\}\) is known as a contraction semigroup.

For example, for a bounded linear operator \(\mathbb B\), define the operator exponential as \(e^{\mathbb Bt}:=\sum_{k=0}^\infty\cfrac{1}{k!}t^k\mathbb B^k\), then it can be shown that \(e^{\mathbb B(t+s)}=e^{\mathbb Bt}e^{\mathbb Bs}\), and from the definition \(e^{\mathbb B0}=I\), hence \(\{e^{\mathbb Bt}:t\geq 0\}\) is a semigroup. As a more concrete example, let \(\mathbb B=\cfrac{d}{dx}:C^\omega \to C^\omega(\mathbb R)\), where \(C^\omega(\mathbb R)\) is the class of analytic functions on the whole of \(\mathbb R\). On \(C^\omega(\mathbb R)\), \(\cfrac{\wrt}{\wrt x}\) is bounded. Then \(e^{\frac{\wrt}{\wrt x} t } f(x) = \sum_{k=0}^\infty \cfrac{1}{k!}t^k\cfrac{\wrt^k}{\wrt x^k}f(x)=f(x-t)\), and the sum converges since \(f\in C^\omega(\mathbb R)\); the series representation is just the Taylor series of \(f\) about the point \(x\).

\subsubsection{Infinitesimal generators} The \emph{infinitesimal generator} (or just \emph{generator} for short) of a semigroup \(\mathbb T(t):L\to L\) is the linear operator defined by 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(T(t)f-I),\]
and the domain, \(\mathcal D(\mathbb B)\) is the subspace of \(L\) for which this limit exists. Note that an essential part of the definition of the generator is its domain. For this reason it is common to denote the generator as the pair \((\mathbb B,\mathcal D(\mathbb B))\). Here we will denote the generator as just \(\mathbb B\) and explicitly state its domain where relevant.

To determine the generator \(\mathbb B\) we can proceed by either differentiation of the semigroup, 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(T(t)f-I),\]
or integration;
\[\int_{t=0}^\infty e^{-st}T(t)\wrt t = (sI-\mathbb B)^{-1}=:R_s.\]
The operator \(R_s\) is known as the resolvent.

% \section{Markov processes and semigroups}
% Semigroups arise naturally in Markov processes. Let \(\{X(t):t\geq 0\}\) be a Markov process with state space \(\mathcal S\). The Markov property states that for \(s,t\geq 0\), \(X(t+s)\) is independent of \(X(u),\, 0\leq u<t\) given \(X(t)\). That is, \(P(X(t+s)\in\mathcal A \mid X(t), X(u), 0\leq u<t)=P(X(t+s)\in\mathcal A \mid X(t))\). If \(P(X(t+s)\in\mathcal A \mid X(t))=P(X(s)\in\mathcal A \mid X(0))\), that is the distributions are invariant under translation by \(t\), then \(X(t)\) is said to be time-homogeneous. Where necessary, we will use the notation \(X_\mu(t)\) to denote the Markov process with initial distribution \(\mu\), i.e.~\(P(X(0)\in\cdot)=\mu(\cdot)\), and \(X_x(t)\) to be the Markov process with initial distribution \(\delta(x)\). Markov processes can be characterised by \emph{probability kernels}.

% Let \((U,\mathcal U)\) and \((V,\mathcal V)\) be two measure spaces, then a probability kernel \(\nu\), is a mapping \(\nu:U\times \mathcal V\to \mathbb R_+\) such that \(\nu(u,\mathcal A)\) is \(\mathcal U\)-measurable in \(u\in U\) for fixed \(\mathcal A\in \mathcal V\), and a probability measure in \(\mathcal A\in \mathcal V\) for fixed \(u\in U\). We say that \(\nu\) is a probability kernel from \(U\) to \(V\).

% With a probability kernel comes associated integral operators. For a measure \(\mu\) on \((U,\mathcal U)\), then 
% \[\mu \nu (\mathcal A)=\int_{U} \mu(\wrt x) \nu(x,\mathcal A)\]
% is also a measure on \((U,\mathcal U)\). For a function \(f:V\to \mathbb R\) then 
% \[\nu f (x)=\int_{V}  \nu(x,\wrt y)f(y).\] 
% We can also compose two transition kernels. Let \(\nu\) be a transition kernel from \(U\) to \(V\) and \(\eta\) be a transition kernel from \(V\) to \(W\). Then \(\nu\eta\) is also a probability kernel from \(U\) to \(W\);
% \[\nu\eta(x,\mathcal A) = \int_{V}\nu(x,\wrt y)\eta(y,\mathcal A).\]

% Let \(\mu_{t}(\cdot)=P(X(t)\in \cdot)\) for \(t\geq 0\) be a one-parameter family of measures on \((\mathcal S,\mathcal F)\), and \(\nu_{t}(x,\cdot)=P(X(t)\in\cdot \mid X(0)=x)\) for \(t\geq 0\) be a one-parameter family of probability kernels from \(\mathcal S\) to \(\mathcal S\). Then \(\mu_t\) and \(\nu_{t}\) characterise the time-homogeneous Markov process \(X(t)\). The probability kernel \(\nu_t\) is referred to as a transition kernel. The kernels \(\nu_t\) form a semigroup since, using the law of total probability and time-homogeneity, 
% \begin{align*}
% \nu_{t+s}(x,\mathcal A)&=P(X(t+s)\in\mathcal A\mid X(0)=x)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(t+s)\in\mathcal A\mid X(t)=y)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(s)\in\mathcal A\mid X(0)=y)
% \\&=\int_{y\in\mathcal S}\nu_t(x,\wrt y)\nu_s(y,\mathcal A)=\nu_t\nu_s(x,\mathcal A).
% \end{align*}
